---
title: "Exercises -- Lecture/Week 02"
format: 
  html:
    toc: true
---

# Lecture 02

Lecture 02 deals with chapters 2 and 3 of the reading. Both chapters have practice exercises, going from *Easy* over *Medium* to *Hard*. Solutions for the book exercises can be found [here](http://xcelab.net/rmpubs/rethinking/rethinking_solutions_2.pdf). In addition, R. McElreath provides (weekly) homework with solutions at the [repository](https://github.com/rmcelreath/stat_rethinking_2023/tree/cdc945ae32f400cc43118ea65aa349767ae9e56c/homework) of the course.

## Book Exercises

### Chapter 2

#### *Easy*

##### 2E1. Which of the expressions below correspond to the statement: the probability of rain on Monday?

- (1) Pr(rain)
- (2) Pr(rain|Monday)
- (3) Pr(Monday|rain)
- (4) Pr(rain, Monday)/ Pr(Monday)

##### 2E2. Which of the following statements corresponds to the expression: Pr(Monday|rain)?

- (1) The probability of rain on Monday.
- (2) The probability of rain, given that it is Monday.
- (3) The probability that it is Monday, given that it is raining.
- (4) The probability that it is Monday and that it is raining.

##### 2E3. Which of the expressions below correspond to the statement: the probability that it is Monday,
given that it is raining?

- (1) Pr(Monday|rain)
- (2) Pr(rain|Monday)
- (3) Pr(rain|Monday) Pr(Monday)
- (4) Pr(rain|Monday) Pr(Monday)/ Pr(rain)
- (5) Pr(Monday|rain) Pr(rain)/ Pr(Monday)

##### 2E4. 
The Bayesian statistician Bruno de Finetti (1906–1985) began his book on probability theory
with the declaration: “PROBABILITY DOES NOT EXIST.” The capitals appeared in the original, so
I imagine de Finetti wanted us to shout this statement. What he meant is that probability is a device
for describing uncertainty from the perspective of an observer with limited knowledge; it has no
objective reality. Discuss the globe tossing example from the chapter, in light of this statement. What
does it mean to say “the probability of water is 0.7”?

#### *Medium*

##### 2M1. Recall the globe tossing model from the chapter. Compute and plot the grid approximate posterior distribution for each of the following sets of observations. In each case, assume a uniform prior for p.

- (1) W, W, W
- (2) W, W, W, L
- (3) L, W, W, L, W, W, W

##### 2M2. 
Now assume a prior for p that is equal to zero when p < 0.5 and is a positive constant when
p ≥ 0.5. Again compute and plot the grid approximate posterior distribution for each of the sets of
observations in the problem just above.

##### 2M3. 
Suppose there are two globes, one for Earth and one for Mars. The Earth globe is 70% covered
in water. The Mars globe is 100% land. Further suppose that one of these globes—you don’t know
which—was tossed in the air and produced a “land” observation. Assume that each globe was equally
likely to be tossed. Show that the posterior probability that the globe was the Earth, conditional on
seeing “land” (Pr(Earth|land)), is 0.23.

##### 2M4. 
Suppose you have a deck with only three cards. Each card has two sides, and each side is either
black or white. One card has two black sides. The second card has one black and one white side. The
third card has two white sides. Now suppose all three cards are placed in a bag and shuffled. Someone
reaches into the bag and pulls out a card and places it flat on a table. A black side is shown facing up,
but you don’t know the color of the side facing down. Show that the probability that the other side is
also black is 2/3. Use the counting method (Section 2 of the chapter) to approach this problem. This
means counting up the ways that each card could produce the observed data (a black side facing up
on the table).

##### 2M5. 
Now suppose there are four cards: B/B, B/W, W/W, and another B/B. Again suppose a card is
drawn from the bag and a black side appears face up. Again calculate the probability that the other
side is black.

##### 2M6. 
Imagine that black ink is heavy, and so cards with black sides are heavier than cards with white
sides. As a result, it’s less likely that a card with black sides is pulled from the bag. So again assume
there are three cards: B/B, B/W, and W/W. After experimenting a number of times, you conclude that
for every way to pull the B/B card from the bag, there are 2 ways to pull the B/W card and 3 ways to
pull the W/W card. Again suppose that a card is pulled and a black side appears face up. Show that
the probability the other side is black is now 0.5. Use the counting method, as before.
2M7. Assume again the original card problem, with a single card showing a black side face up. Before
looking at the other side, we draw another card from the bag and lay it face up on the table. The face
that is shown on the new card is white. Show that the probability that the first card, the one showing
a black side, has black on its other side is now 0.75. Use the counting method, if you can. Hint: Treat
this like the sequence of globe tosses, counting all the ways to see each observation, for each possible
first card.

#### *Hard*

##### 2H1. 
Suppose there are two species of panda bear. Both are equally common in the wild and live
in the same places. They look exactly alike and eat the same food, and there is yet no genetic assay
capable of telling them apart. They differ however in their family sizes. Species A gives birth to twins
10% of the time, otherwise birthing a single infant. Species B births twins 20% of the time, otherwise
birthing singleton infants. Assume these numbers are known with certainty, from many years of field
research.
Now suppose you are managing a captive panda breeding program. You have a new female panda
of unknown species, and she has just given birth to twins. What is the probability that her next birth
will also be twins?

##### 2H2. 
Recall all the facts from the problem above. Now compute the probability that the panda we
have is from species A, assuming we have observed only the first birth and that it was twins.

##### 2H3. 
Continuing on from the previous problem, suppose the same panda mother has a second birth
and that it is not twins, but a singleton infant. Compute the posterior probability that this panda is
species A.

##### 2H4. 
A common boast of Bayesian statisticians is that Bayesian inference makes it easy to use all of
the data, even if the data are of different types.
So suppose now that a veterinarian comes along who has a new genetic test that she claims can
identify the species of our mother panda. But the test, like all tests, is imperfect. This is the information
you have about the test:

- The probability it correctly identifies a species A panda is 0.8.
- The probability it correctly identifies a species B panda is 0.65.

The vet administers the test to your panda and tells you that the test is positive for species A. First
ignore your previous information from the births and compute the posterior probability that your
panda is species A. Then redo your calculation, now using the birth data as well.

### Chapter 3

#### *Easy*

These problems use the samples from the posterior distribution for the globe tossing example.
This code will give you a specific set of samples, so that you can check your answers exactly.

```{r}
p_grid <- seq( from=0 , to=1 , length.out=1000 )
prior <- rep( 1 , 1000 )
likelihood <- dbinom( 6 , size=9 , prob=p_grid )
posterior <- likelihood * prior
posterior <- posterior / sum(posterior)
set.seed(100)
samples <- sample( p_grid , prob=posterior , size=1e4 , replace=TRUE )
```

Use the values in `samples` to answer the questions that follow.

3E1. How much posterior probability lies below p = 0.2?

3E2. How much posterior probability lies above p = 0.8?

3E3. How much posterior probability lies between p = 0.2 and p = 0.8?
3E4. 20% of the posterior probability lies below which value of p?

3E5. 20% of the posterior probability lies above which value of p?

3E6. Which values of p contain the narrowest interval equal to 66% of the posterior probability?

3E7. Which values of p contain 66% of the posterior probability, assuming equal posterior probability
both below and above the interval?

#### *Medium*

3M1. Suppose the globe tossing data had turned out to be 8 water in 15 tosses. Construct the posterior
distribution, using grid approximation. Use the same flat prior as before.

3M2. Draw 10,000 samples from the grid approximation from above. Then use the samples to calculate
the 90% HPDI for p.

3M3. Construct a posterior predictive check for this model and data. This means simulate the distribution
of samples, averaging over the posterior uncertainty in p. What is the probability of observing
8 water in 15 tosses?

3M4. Using the posterior distribution constructed from the new (8/15) data, now calculate the probability
of observing 6 water in 9 tosses.

3M5. Start over at 3M1, but now use a prior that is zero below p = 0.5 and a constant above p = 0.5.
This corresponds to prior information that a majority of the Earth’s surface is water. Repeat each
problem above and compare the inferences. What difference does the better prior make? If it helps,
compare inferences (using both priors) to the true value p = 0.7.

3M6. Suppose you want to estimate the Earth’s proportion of water very precisely. Specifically, you
want the 99% percentile interval of the posterior distribution of p to be only 0.05 wide. This means
the distance between the upper and lower bound of the interval should be 0.05. How many times will
you have to toss the globe to do this?

#### *Hard*

Introduction. The practice problems here all use the data below. These data indicate the gender
(male=1, female=0) of officially reported first and second born children in 100 two-child families.

```{r}
birth1 <- c(1,0,0,0,1,1,0,1,0,1,0,0,1,1,0,1,1,0,0,0,1,0,0,0,1,0,
0,0,0,1,1,1,0,1,0,1,1,1,0,1,0,1,1,0,1,0,0,1,1,0,1,0,0,0,0,0,0,0,
1,1,0,1,0,0,1,0,0,0,1,0,0,1,1,1,1,0,1,0,1,1,1,1,1,0,0,1,0,1,1,0,
1,0,1,1,1,0,1,1,1,1)
birth2 <- c(0,1,0,1,0,1,1,1,0,0,1,1,1,1,1,0,0,1,1,1,0,0,1,1,1,0,
1,1,1,0,1,1,1,0,1,0,0,1,1,1,1,0,0,1,0,1,1,1,1,1,1,1,1,1,1,1,1,1,
1,1,1,0,1,1,0,1,1,0,1,1,1,0,0,0,0,0,0,1,0,0,0,1,1,0,0,1,0,0,1,1,
0,0,0,1,1,1,0,0,0,0)
```


So for example, the first family in the data reported a boy (1) and then a girl (0). The second family
reported a girl (0) and then a boy (1). The third family reported two girls. You can load these two
vectors into R’s memory by typing:

```{r}
library(rethinking)
data(homeworkch3)
```

Use these vectors as data. So for example to compute the total number of boys born across all of these
births, you could use:

```{r}
sum(birth1) + sum(birth2)
```

3H1. Using grid approximation, compute the posterior distribution for the probability of a birth
being a boy. Assume a uniform prior probability. Which parameter value maximizes the posterior
probability?

3H2. Using the sample function, draw 10,000 random parameter values from the posterior distribution
you calculated above. Use these samples to estimate the 50%, 89%, and 97% highest posterior
density intervals.

3H3. Use rbinom to simulate 10,000 replicates of 200 births. You should end up with 10,000 numbers,
each one a count of boys out of 200 births. Compare the distribution of predicted numbers
of boys to the actual count in the data (111 boys out of 200 births). There are many good ways to
visualize the simulations, but the dens command (part of the rethinking package) is probably the
easiest way in this case. Does it look like the model fits the data well? That is, does the distribution
of predictions include the actual observation as a central, likely outcome?

3H4. Now compare 10,000 counts of boys from 100 simulated first borns only to the number of boys
in the first births, birth1. How does the model look in this light?

3H5. The model assumes that sex of first and second births are independent. To check this assumption,
focus now on second births that followed female first borns. Compare 10,000 simulated counts
of boys to only those second births that followed girls. To do this correctly, you need to count the
number of first borns who were girls and simulate that many births, 10,000 times. Compare the
counts of boys in your simulations to the actual observed count of boys following girls. How does the
model look in this light? Any guesses what is going on in these data?

---

## Homework

### Week 01 (Compromises of Lecture 01 and Lecture 02)

[Exercise](https://github.com/rmcelreath/stat_rethinking_2023/blob/cdc945ae32f400cc43118ea65aa349767ae9e56c/homework/week01.pdf)

[Solution](https://github.com/rmcelreath/stat_rethinking_2023/blob/cdc945ae32f400cc43118ea65aa349767ae9e56c/homework/week01_solutions.pdf)



